\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-06 manuscript.Rmd
manu

---
title             : "**The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates**"
shorttitle        : "Illusion Game"
author:
  - name          : "Dominique Makowski"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dom.makowski@gmail.com"
    orcid         : 0000-0001-5375-9967

affiliation:
  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Abstract abstract. We invite researchers to re-analyze our open-access data to provide complimentary evidence (or absence thereof) on the effect and structure of illusion sensitivity.

@@

Notably, the presence of a common mechanism underlying the effect of different illusions has been contested [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]; and the nature of the underlying processes - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] are strongly debated. The existence of dispositional or demographic correlates of illusion sensitivity - for example, higher illusion resistance has been reported in schizophrenia and autism [@notredame2014; @park2022; @giaouri2011; @keane2014]; as well as in males as compared to females [@miller2001; @papageorgiou2020; @lo2011; @shaqiri2018] - is another area of controversy.

@@

```{r message=FALSE, warning=FALSE, fig.cap="The parametric framework for visual illusions (Makowski et al., 2021) applied to the Müller-Lyer illusion (above). Below are examples of stimuli showcasing the manipulation of two parameters, task difficulty and illusion strength."}
knitr::include_graphics("figures/Figure1.png")
```

@@

Indeed, many visual illusions can be seen as being composed of *targets* (e.g., same-length lines), of which perception is biased by the *context* (e.g., in the Müller-Lyer illusion, the same-length line segments appear to have different lengths when they end with inwards or outwards pointing arrows). 
Past illusion studies traditionally employed paradigms focusing on participants' subjective experience, by asking them to what extent they perceive two identical targets as different [@lányi], or having them adjust the targets to a reference stimulus relying only on their perception [@mylniec2016a; @grzeczkowski2018]. 
Alternatively, *Pyllusion* allows the creation of illusions in which the targets are objectively different (e.g., one segment is truly more or less longer than the other), and in which the illusion varies in strength (the biasing angle of the arrows is more or less acute).

@@

In the present set of preregistered studies, we will first test this novel paradigm by investigating if the effect of illusion and task difficulty can be manipulated continuously, and separated statistically. Then, we will further utilize the paradigm to assess whether 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) share a common latent factor. Finally, we will investigate how the the inter-individual sensitivity to illusions relates to dispositional variables, such as demographic characteristics and personality.

@@

Using the GAMs as a "ground-truth", we attempted at approximating them using general linear models, which have the advantage of estimating the participant-level variability of the effects (via random slopes). Following a comparison of models with a combination of transformations (raw, log, square root or cubic root) on the main predictors (task *difficulty* and illusion *strength*), we selected and fitted the best model (best on their indices of fit), and compared their output visually (see **Figure 2**).

@@

The effect on errors was monotonic for most illusions, with the exception of Delboeuf and Zöllner. For both of them, mildly congruent illusion strengths (which theoretically were supposed to be associated with less errors than incongruent effects) were related to small and strong increases of errors, respectively. For the Delboeuf illusion, we believe that this was due to an artifact caused by the illusion generation algorithm: the outline of the target circles was always created as slightly bigger, which made the difference between them more obvious at an illusion strength of 0. This was fixed in latest release of *Pyllusion* (v1.2), which now generate outlines of the same size as the target circle. For the Zöllner illusion, this observation is consistent with @kitaoka2000, in which they reported an acute angle contraction effect at very small as well as at sufficiently large angles (below 10 degrees for the former and between 50 to 90 degrees for the latter) between the target horizontal line and the biasing horizontal bars when the illusion strength is weak. This invoked underlying acute angle contraction mechanism could potentially explain the non-monotonic pattern found [@kitaoka2007; @kitaoka2000a].

Finally, this study provided evidence for both the existence of illusion-specific factors, as well as for a common latent factor (labelled Factor *i*) that explained about half of the total variance. These participant-level scores were related to the error rate and average reaction time, and can thus be interpreted as indices of illusion sensitivity.

@@

Study 3 aimed at investigating the links between the inter-individual scores of illusion sensitivity (obtained in study 2), and demographic and dispositional variables. **Insert a bit of litterature about why and some findings justifying what we did**.

@@

```{r message=FALSE, warning=FALSE, fig.cap="CAPTION."}
knitr::include_graphics("figures/Figure3.png")
```

>>>>

---
title             : "**The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates**"
shorttitle        : "Illusion Game"
author:
  - name          : "Dominique Makowski"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "HSS 04-18, 48 Nanyang Avenue, Singapore"
    email         : "dom.makowski@gmail.com"
    orcid         : 0000-0001-5375-9967
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Data curation"
      - "Formal Analysis"
      - "Funding acquisition"
      - "Investigation"
      - "Methodology"
      - "Project administration"
      - "Resources"
      - "Software"
      - "Supervision"
      - "Validation"
      - "Visualization"
      - "Writing – original draft"
  - name          : "An Shu Te"
    affiliation   : "1"
    orcid         : 0000-0001-8008-2824
    role:
      - "Project administration"
      - "Resources"
      - "Investigation"
      - "Writing – original draft"
  - name          : "Stephanie Kirk"
    affiliation   : "1"
    orcid         : 0000-0001-8008-2824
    role:
      - "Project administration"
      - "Resources"
      - "Writing – original draft"
  - name          : "Ngoi Zi Liang"
    affiliation   : "1"
    orcid         : 0000-0001-8008-2824
    role:
      - "Project administration"
      - "Resources"
      - "Writing – original draft"
  - name          : "S.H. Annabel Chen"
    affiliation   : "1, 2, 3, 4"
    orcid         : 0000-0002-1540-5516
    role:
      - "Project administration"
      - "Supervision"
      - "Writing – review & editing"

affiliation:
  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"

authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Abstract abstract. We invite researchers to re-analyze our open-access data to provide complimentary evidence (or absence thereof) on the effect and structure of illusion sensitivity.

@@

Notably, the presence of a common mechanism underlying the effect of different illusions has been contested [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]; and the nature of the underlying processes - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] are strongly debated. The existence of dispositional or demographic correlates of illusion sensitivity - for example, higher illusion resistance has been reported in schizophrenia and autism [@notredame2014; @park2022; @giaouri2011; @keane2014]; as well as in males as compared to females [@miller2001; @papageorgiou2020; @lo2011; @shaqiri2018] (**Is there something for personality that we could use here instead of sex?**) - is another area of controversy.

@@

```{r message=FALSE, warning=FALSE, fig.cap="The parametric framework for visual illusions (Makowski et al., 2021) applied to the Müller-Lyer illusion (above). Below are examples of stimuli showcasing the manipulation of two parameters, task difficulty and illusion strength.", out.width="100%"}
knitr::include_graphics("figures/Figure1.png")
```

@@

Indeed, many visual illusions can be seen as being composed of *targets* (e.g., same-length lines), of which perception is biased by the *context* (e.g., in the Müller-Lyer illusion, the same-length line segments appear to have different lengths when they end with inwards or outwards pointing arrows). 
Past illusion studies traditionally employed paradigms focusing on participants' subjective experience, by asking them to what extent they perceive two identical targets as different [@lanyi2022can], or having them adjust the targets to a reference stimulus relying only on their perception [@mylniec2016a; @grzeczkowski2018]. 
Alternatively, *Pyllusion* allows the creation of illusions in which the targets are objectively different (e.g., one segment is truly more or less longer than the other), and in which the illusion varies in strength (the biasing angle of the arrows is more or less acute).

@@

In the present set of preregistered studies, we will first test this novel paradigm by investigating if the effect of illusion and task difficulty can be manipulated continuously, and separately modeled statistically. Then, we will further utilize the paradigm to assess whether 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) share a common latent factor. Finally, we will investigate how the the inter-individual sensitivity to illusions relates to dispositional variables, such as demographic characteristics and personality.

@@

Using GAMs as the "ground-truth" models, we attempted at approximating them using general linear models, which have the advantage of estimating the participant-level variability of the effects (via random slopes). Following a comparison of models with a combination of transformations (raw, log, square root or cubic root) on the main predictors (task *difficulty* and illusion *strength*), we selected and fitted the best model (best on their indices of fit), and compared their output visually (see **Figure 2**).

@@

The effect on errors was monotonic for most illusions, with the exception of Delboeuf and Zöllner. For both of them, mildly congruent illusion strengths (which theoretically were supposed to be associated with less errors than incongruent effects) were related to small and strong increases of errors, respectively. For the Delboeuf illusion, we believe that this was due to an artifact caused by the illusion generation algorithm: the outline of the target circles was always created as slightly bigger, which made the difference between them more obvious at an illusion strength of 0. This was fixed in latest release of *Pyllusion* (v1.2), which now generate outlines of the same size as the target circle. For the Zöllner illusion, the observed non-monotonic pattern is actually consistent with previous reports [@kitaoka2000; @kitaoka2007], suggesting an acute angle contraction effect at very small as well as at sufficiently large angles (below 10 degrees for the former and between 50 to 90 degrees for the latter) between the target horizontal line and the biasing horizontal bars when the illusion strength is weak.

Finally, this study provided evidence for both the existence of illusion-specific factors, as well as for a common latent factor (labelled Factor *i*) that explained about half of the total variance. These participant-level scores were positively related to the error rate and average reaction time, and can thus be interpreted as indices of illusion sensitivity.

@@

Study 3 aimed at investigating the links between the inter-individual scores of illusion sensitivity (obtained in study 2), and demographic and dispositional variables. **TODO: Insert a bit of literature about why and some findings justifying what we did**.

@@

```{r message=FALSE, warning=FALSE, fig.cap="The upper plots show the distribution of demographic and dispositional variables. The middle plots show the relationship between illusion sensitivty scores, sex, and age (solid lines indicate significant relationships). Bottom plots show the correlation between the general factor of illusion sensitivity (Factor \\textit{i}), and personality traits.", out.width="100%"}
knitr::include_graphics("figures/Figure3.png")
```
